Xolotl: An Intuitive and Comprehensible Neuronal Simulator

### 2018-05-23 PaperLatex/frontiers.tex
Begin simulation section

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage{natbib}
\usepackage{placeins}

\linenumbers

@@

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\texttt{xolotl} is an open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Complex models and networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models can be specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. It also includes a customizable graphical user interface (`puppeteer`) for rapid prototyping and hand-tuning conductances in real-time. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. \texttt{xolotl} is freely available at \url{https://github.com/marderlab/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

@@

\texttt{xolotl} comes packaged with visualization functions, a graphical user interface for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in \texttt{xolotl} methods. The \texttt{puppeteer} toolbox permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of complex neural network models, facilitate collaborative modeling, and complement other tools being developed in the neuroinformatics community.

@@

@@

Once the \texttt{xolotl} object is instantiated in \texttt{MATLAB}, the \texttt{add} function constructs new \texttt{cpplab} objects. A model neuron consists of one or more compartments, each representing a section of membrane with capacitance and surface area. Isopotential models require one compartment. Models with multiple neurons, units, or non-trivial morphology require multiple compartments. All specifiable properties of compartments are shown in Supplementary Table 1.

\texttt{xolotl} provides some features for generating complex models. Synapses can be added with the \texttt{connect} function. At minimum possess identifiers to presynaptic and postsynaptic compartments and default to electrical synapses. All specifiable properties of synapses are shown in Supplementary Table 2. To create axons or transport chains, the \texttt{slice} function splits a compartment into $n$ discrete segments and adds these compartments to the network connected by electrical synapses.

>>>>

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage{natbib}
\usepackage{placeins}
\usepackage{matlab-prettifier}
\linenumbers

@@

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\texttt{xolotl} is an open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Complex models and networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models can be specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. It also includes a customizable graphical user interface (GUI, `puppeteer`) for rapid prototyping and hand-tuning conductances in real-time. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. \texttt{xolotl} is freely available at \url{https://github.com/marderlab/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

@@

\texttt{xolotl} comes packaged with visualization functions, a graphical user interface (GUI) for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in \texttt{xolotl} methods. The \texttt{puppeteer} toolbox permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of complex neural network models, facilitate collaborative modeling, and complement other tools being developed in the neuroinformatics community.

@@

\subsubsection{Using the \texttt{cpplab} Framework}

The \texttt{add} function will construct a \texttt{cpplab} object and affix it to as a field in the \texttt{xolotl} structure. All compartments, conductances, synapses, and controllers are \texttt{cpplab}. Compartments add to the \texttt{xolotl} object and conductances add to compartments. Specific properties can be specified using key-value pair arguments (e.g. \ref{fig:figurehh}A).

\texttt{cpplab} comes with several features which simplify the handling of high-dimensional models. The \texttt{find} function acquires a cell array of all properties of the network which satisfy a search condition.

@@

A model neuron consists of one or more compartments, each representing a section of membrane with capacitance and surface area. Isopotential models require one compartment, whereas models with multiple neurons, units, or non-trivial morphology require multiple compartments. All specifiable properties of compartments are shown in Supplementary Table 1.

\texttt{xolotl} provides some features for generating complex models. Synapses can be added with the \texttt{connect} function. At minimum synapses possess identifiers to presynaptic and postsynaptic compartments and default to electrical synapses. All specifiable properties of synapses are shown in Supplementary Table 2. To create axons or transport chains, the \texttt{slice} function splits a compartment into $n$ discrete segments and adds these compartments to the network connected by electrical synapses.

\subsubsection{Conductances and Controllers}

All conductances contain fields for maximal conductance and reversal potential. Conductances with activation and inactivation variables include them as $m$ and $h$ respectively. Gating functions and their respective time constants are contained within the conductance header file. \texttt{xolotl} comes packaged with conductances from several dozen papers (Supplementary Table 3).

\subsubsection{Creating Custom \texttt{cpplab} Objects}

\texttt{xolotl} contains template header files for producing custom conductances. The template contains instructions on how to design novel conductances with arbitrary specifications.

\subsection{Simulation}

Models are simulated in \texttt{xolotl} with the \texttt{integrate} function which outputs as time series the membrane potentials, intracellular calcium concentrations, controller states, intrinsic currents, and synaptic currents. The \texttt{integrate} function also accepts an argument which specifies injected current or clamped voltage.

\texttt{xolotl} uses the exponential Euler method (Dayan \& Abbott 2001) for single compartment models, forward Euler for gating variables, and a Crank-Nicholson regime for electrically-coupled compartments. The simulation time-resolution can be specified to target arbitrary precision, and an output time step can be selected to support automatic down-sampling for memory considerations.

Simulations can be run in "open-loop" mode where each simulation begins by resetting all dynamical variables to their initial conditions at instantiation, or "closed-loop" mode which begins simulation with the current network state.

\subsection{Using the \texttt{puppeteer} graphical interface}

\texttt{xolotl} comes packaged with a graphical user interface for visualizing parameter changes in real-time. The \texttt{manipulate} function begins the
